Data Prep

### 2024-02-21 scripts/00.0_dataPrep.qmd
restricted to southern individuals for both movement and social

@@

# Calculate daily modes and add them to the dataset (using the function defined above)
with_modes <- map(seasons_list, ~.x %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(diff = as.numeric(difftime(lead(timestamp), timestamp, units = "mins"))) %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(mode = Mode(round(diff))) %>%
                 ungroup())

@@

# Tidy up
rm(seasons_list)
rm(season_names)
rm(low_fix_rate_indivs)
rm(with_modes)
gc()
```

@@

```{r}
load(here::here("data", "dataPrep", "seasons_list.Rda")) 
load(here::here("data", "dataPrep", "low_fix_rate_indivs.Rda"))
```

```{r}
removed_lfr <- map2(seasons_list, low_fix_rate_indivs,                     ~.x %>% filter(!(Nili_id %in% .y))) 
save(removed_lfr, file = here::here("data", "dataPrep", "removed_lfr.Rda"))  
# tidy up 
rm(seasons_list) 
rm(low_fix_rate_indivs) 
rm(removed_lfr) 
gc()
```

@@

@@

```{r}
#| eval: false
# Transform to SF object, so we can get centroids
seasons_sf <- map(removed_lfr, ~.x %>%
                   sf::st_as_sf(coords = c("location_long", "location_lat"), remove = F) %>%
                   sf::st_set_crs("WGS84") %>%
                   sf::st_transform(32636))

## Examine a histogram of centroid latitudes 
walk(centroids, ~hist(st_coordinates(.x)[,2])) # looks like 3550000 is generally a good cutoff point here.

## Remove individuals not in the south
removed_northern <- map2(.x = removed_lfr, 
                         .y = southern_indivs, ~.x 
                         %>% filter(Nili_id %in% .y))
save(removed_northern, file = here::here("data", "dataPrep", "removed_northern.Rda"))

# tidy up
rm(removed_lfr)
rm(season_names)
rm(seasons_sf)
rm(centroids)
rm(southern_indivs)
rm(removed_northern)
rm(df)
gc()
```

The next step will be to remove individuals with too few points per day or too few days tracked. But we don't want those indivs removed for the *social* analysis, since those things will be accounted for with SRI and all individuals make up important parts of the social fabric. So when we do the social analysis, we will use `seasons_list` and not anything further along.

```{r}
load(here::here("data", "dataPrep", "removed_northern.Rda"))
```

```{r}
#| eval: false
# Get roosts for each season ----------------------------------------------
roosts <- furrr::future_map(removed_northern, ~vultureUtils::get_roosts_df(df = .x, id = "Nili_id"), .progress = T)
roosts <- roosts %>%
  furrr::future_map(., ~st_as_sf(.x, crs = "WGS84", 
                   coords = c("location_long", "location_lat"), 
                   remove = F), .progress = T)
save(roosts, file = here::here("data", "dataPrep", "roosts.Rda"))

# Tidy up
rm(removed_northern)
rm(roosts)
gc()
```

```{r}
#| eval: false
# Remove nighttime points -------------------------------------------------
removed_nighttime <- map(removed_northern, ~{
  times <- suncalc::getSunlightTimes(date = unique(lubridate::date(.x$timestamp)),
                                     lat = 31.434306, lon = 34.991889, # XXX revisit this!
                                     keep = c("sunrise", "sunset")) %>%
    dplyr::select("dateOnly" = date, sunrise, sunset)

  .x <- .x %>%
    dplyr::mutate(dateOnly = lubridate::ymd(dateOnly)) %>%
    dplyr::left_join(times, by = "dateOnly") %>%
    dplyr::mutate(daylight = ifelse(timestamp >= sunrise & timestamp <= sunset, "day", "night")) %>%
    dplyr::filter(daylight == "day")
})

@@

# tidy up
rm(removed_nighttime)
rm(removed_northern)
rm(season_names)
gc()
```

@@

@@

```{r}
load(here::here("data", "dataPrep", "removed_northern.Rda"))
```

```{r}
#| eval: false
# Calculate number of points per day for each individual
ppd <- map_dfr(removed_northern, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T)))

@@

## Keep days with more than 10 points, or with fewer than 10 if the minimum battery charge that day is >50%
removed_lowppd <- map(removed_northern, ~.x %>%
                 group_by(Nili_id, dateOnly) %>%
                 filter(n() >= 10 | (n() < 10 & 
                                       min(battery_charge_percent, 
                                           na.rm = T) > 50)) %>%
                 ungroup())

@@

# Include only individuals with enough days tracked -----------------------
indivsToKeep <- map(removed_lowppd, ~.x %>%
                       st_drop_geometry() %>%
                       group_by(Nili_id) %>%
                       summarize(nDaysTracked = 
                                   length(unique(dateOnly))) %>%
                       filter(nDaysTracked >= 30) %>%
                       pull(Nili_id))

@@

beforeIndivs <- map_dbl(removed_lowppd, ~length(unique(.x$Nili_id)))
afterIndivs <- map_dbl(removed_too_few_days, ~length(unique(.x$Nili_id)))

@@

# tidy up
rm(removed_lowppd)
rm(season_names)
rm(season_durations)
rm(indivsToKeep)
rm(removed_too_few_days)
rm(beforeIndivs)
rm(afterIndivs)
gc()
```

@@

```{r}
#| eval: false
# Get elevation rasters ---------------------------------------------------
sf <- map(removed_too_few_days, ~st_as_sf(.x, coords = c("location_lat", "location_long"), remove = F, crs = "WGS84"))
elevs_z10 <- map(sf, ~elevatr::get_elev_raster(.x , z = 10))
save(elevs_z10, file = here::here("data", "dataPrep", "elevs_z10.Rda"))
save(sf, file = here::here("data", "dataPrep", "sf.Rda"))
```

@@

```{r}
#| eval: false
groundElev_z10 <- map2(elevs_z10, sf, ~raster::extract(x = .x, y = .y)) # have to convert the crs because the raster from elevatr is in WGS84.

>>>>

```{r}
#| eval: false
# Transform to SF object, so we can get centroids
seasons_sf <- map(seasons_list, ~.x %>%
                   sf::st_as_sf(coords = c("location_long", "location_lat"), remove = F) %>%
                   sf::st_set_crs("WGS84") %>%
                   sf::st_transform(32636))

## Examine a histogram of centroid latitudes 
#walk(centroids, ~hist(st_coordinates(.x)[,2])) # looks like 3550000 is generally a good cutoff point here.

## Remove individuals not in the south
removed_northern <- map2(.x = seasons_list, 
                         .y = southern_indivs, ~.x 
                         %>% filter(Nili_id %in% .y))
save(removed_northern, file = here::here("data", "dataPrep", "removed_northern.Rda"))

# tidy up
rm(seasons_list)
rm(season_names)
rm(seasons_sf)
rm(centroids)
rm(southern_indivs)
rm(removed_northern)
rm(df)
gc()
```

@@

# Calculate daily modes and add them to the dataset (using the function defined above)
with_modes <- map(removed_northern, ~.x %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(diff = as.numeric(difftime(lead(timestamp), timestamp, units = "mins"))) %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(mode = Mode(round(diff))) %>%
                 ungroup())

@@

# Tidy up
rm(removed_northern)
rm(season_names)
rm(low_fix_rate_indivs)
rm(with_modes)
gc()
```

@@

```{r}
load(here::here("data", "dataPrep", "removed_northern.Rda")) 
load(here::here("data", "dataPrep", "low_fix_rate_indivs.Rda"))
```

```{r}
removed_lfr <- map2(removed_northern, low_fix_rate_indivs,                     ~.x %>% filter(!(Nili_id %in% .y))) 
save(removed_lfr, file = here::here("data", "dataPrep", "removed_lfr.Rda"))  
# tidy up 
rm(removed_northern) 
rm(low_fix_rate_indivs) 
rm(removed_lfr) 
gc()
```

@@

The next step will be to remove individuals with too few points per day or too few days tracked. But we don't want those indivs removed for the *social* analysis, since those things will be accounted for with SRI and all individuals make up important parts of the social fabric. So when we do the social analysis, we will use `removed_lfr` and not anything further along.

-   `data/dataPrep/downsampled_10min_forSocial.Rda`

```{r}
load(here::here("data", "dataPrep", "downsampled_10min_forSocial.Rda"))
```

```{r}
#| eval: false
# Get roosts for each season ----------------------------------------------
roosts <- map(downsampled_10min_forSocial, ~vultureUtils::get_roosts_df(df = .x, id = "Nili_id"), .progress = T)
roosts <- roosts %>%
  map(., ~st_as_sf(.x, crs = "WGS84", 
                   coords = c("location_long", "location_lat"), 
                   remove = F), .progress = T)
save(roosts, file = here::here("data", "dataPrep", "roosts.Rda"))

# Tidy up
rm(downsampled_10min_forSocial)
rm(roosts)
gc()
```

@@

```{r}
#| eval: false
# Remove nighttime points -------------------------------------------------
removed_nighttime <- map(removed_lfr, ~{
  times <- suncalc::getSunlightTimes(date = unique(lubridate::date(.x$timestamp)),
                                     lat = 31.434306, lon = 34.991889, # XXX revisit this!
                                     keep = c("sunrise", "sunset")) %>%
    dplyr::select("dateOnly" = date, sunrise, sunset)

  .x <- .x %>%
    dplyr::mutate(dateOnly = lubridate::ymd(dateOnly)) %>%
    dplyr::left_join(times, by = "dateOnly") %>%
    dplyr::mutate(daylight = ifelse(timestamp >= sunrise & timestamp <= sunset, "day", "night")) %>%
    dplyr::filter(daylight == "day")
}, .progress = T)

@@

# tidy up
rm(removed_nighttime)
rm(removed_lfr)
rm(season_names)
gc()
```

@@

-   `data/dataPrep/removed_nighttime.Rda`

@@

```{r}
load(here::here("data", "dataPrep", "removed_nighttime.Rda"))
```

```{r}
#| eval: false
# Calculate number of points per day for each individual
ppd <- map_dfr(removed_nighttime, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T)), .progress = T)

@@

## Keep days with more than 10 points, or with fewer than 10 if the minimum battery charge that day is >50%
removed_lowppd <- map(removed_nighttime, ~.x %>%
                 group_by(Nili_id, dateOnly) %>%
                 filter(n() >= 10 | (n() < 10 & 
                                       min(battery_charge_percent, 
                                           na.rm = T) > 50)) %>%
                 ungroup())

@@

# Include only individuals with enough days tracked -----------------------
indivsToKeep <- map(removed_lowppd, ~.x %>%
                       st_drop_geometry() %>%
                       group_by(Nili_id) %>%
                       summarize(nDaysTracked = 
                                   length(unique(dateOnly))) %>%
                       filter(nDaysTracked >= 30) %>%
                       pull(Nili_id), .progress = T)

@@

@@

# tidy up
rm(removed_lowppd)
rm(season_names)
rm(season_durations)
rm(indivsToKeep)
rm(removed_too_few_days)
gc()
```

@@

```{r}
#| eval: false
# Get elevation rasters ---------------------------------------------------
sf <- map(removed_too_few_days, ~st_as_sf(.x, coords = c("location_lat", "location_long"), remove = F, crs = "WGS84"))
elevs_z10 <- map(sf, ~elevatr::get_elev_raster(.x , z = 10), .progress = T)
save(elevs_z10, file = here::here("data", "dataPrep", "elevs_z10.Rda"))
save(sf, file = here::here("data", "dataPrep", "sf.Rda"))
```

@@

```{r}
#| eval: false
groundElev_z10 <- map2(elevs_z10, sf, ~raster::extract(x = .x, y = .y), .progress = T) # have to convert the crs because the raster from elevatr is in WGS84.
